---
title: "Assignment 3"
author: "Group 4"
output: pdf_document
---

# Instructions

Consider the stochastic process of spread of an epidemics (of COVID-19, influenza, or other similar viruses) in a population composed by N individuals.
Assume that each individual at each time step can be in one of the following 3 states:
- S = susceptible = not infected and not immune (then s/he can be infected)
- I = infected
- R = removed = recovered and immune, or dead

Assume also that:
- at each time step each susceptible individual becomes infected with probability 0.08
- each infected individual is removed with probability 0.6
- each removed individual returns back to be susceptible with probability 0.1.

Represent the evolution of the epidemics as a Markov chain and write the transition matrix.

Imagine now that the number N of individuals in the population is N = 10000.

If is it possible, find what will be the number of individuals in the three states S,I,R in the long run, after many time steps.

# Theoretical Background



# Solution
We can approach this kind of problem as an oriented graph where:
- nodes are the states
- edges are the transitions between a state and another

```{r The package for work with Markow Chain is downloaded as follows}
library("markovchain")
```


being a stochastic markov chain:
- all the elements are >= 0
- the sum of rows and columns are = 1
then we can infer which are the remaining elements of the transition matrix.

```{r Initializing variables}
N <- 10000 # where N is the numeber of individuals
eipdemics_states <- c("S", "I", "R")

# assuming as initial state the start of the epidemics phenomenon
initial_state <- c(N, 0, 0) 

# we instantiate the transition matrix directly as a markov chain data type
by_row <- TRUE

mcEpidemics <- new("markovchain", 
                   states = eipdemics_states,
                   transitionMatrix = matrix(
                     data =  c(0.9, 0.08, 0.02,
                               0.0, 0.4, 0.6,
                               0.1, 0.54, 0.36),
                     byrow = by_row,
                     ncol = 3,
                     nrow = 3),
                   name = "Epidemics")
# return
mcEpidemics
```

As we have just one transition matrix, then the markov chain is homogeneous.
It is easy to see that the chain is irreducible since all the states communicate (it is made by one communicating class only).


```{r Obtain the transition matrix as a column vector}
mcDf <- as(mcEpidemics, "data.frame")
mcMatrix <- as(mcEpidemics, "matrix")
mcNew <- as(mcDf, "markovchain")
mcDf


plot(mcEpidemics)
```

```{r To visualize with a graph the transition matrix}
library(stats)
mcIgraph <- as(mcEpidemics, "igraph")
mcIgraph
```

```{r Steady state of transition matrix}
steadyStates(mcEpidemics)
```



```{r How many transitions do the system need to reach the steady state?}
ans <- initial_state

step <- 0
while (round(ans, digits = 0) != steadyStates(mcEpidemics)) { # we round values as discrete individuals are meaningless in percentage
  ans <- ans %*% mcMatrix # need matrix data type to perform %*% operator
  step <- step + 1
}

print(quote = FALSE, paste0("we reach the steady state in number of steps: ", step))
```



